Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-10-13 draft_lm1.tex
moving Rt estimation to main text

@@

\subsection*{The data}

The data used to fit the model consists of weekly incidence from 1996 to 2014 
in the city of Rio de Janeiro, Brazil.
From this series, we calculated the effective reproduction number, $R_t$, 
according to the expression given by Nishiura et al.~\cite{nishiura}.
The method assumes the disease counts $Y(t)$ are Poisson random variables such 
that $R_t$ is 
\begin{equation}
 \label{eq:Rt}
 R_t = \left(\frac{Y(t + 1)}{Y(t)}\right)^{1/n}
\end{equation}

where $n$ is the ratio between the length of reporting interval and the mean 
generation time of the disease.
Here we obtain the estimates of $R_t$ using a moving window of three weeks, 
which roughly corresponds to the generation time of dengue, and thus $n = 1$.

We provide more detail on drawing posterior inference for $R_t$ in Text S1.

@@

As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number as defined in~\ref{eq:Rt}.
We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
 \frac{dS}{dt} &= -\beta(t)SI \\
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\
 \frac{dR}{dt} &= \tau I&
\end{align}
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
This is a rather simplified model, in which, for instance, the vector is left 
out.
Also, although there probably are multiple circulating serotypes, our approach
does not discriminate between them due to the lack of serotype-specific data.

@@

For the fitting procedure the time series was divided into $J=10$ epidemic 
windows that corresponded to significant raises in incidence and normalized to 
lie on the $[0,1]$ interval.

Also, as the numerical integration routine needs to treat $\beta(t)$ as a
continuous function of time because of the variable step size, we used linear 
interpolation to obtain values of $R_t$ for any time point.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techiniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at 
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubins' R convergence diagnostic~\cite{brooks1998}.
% Results and Discussion can be combined.
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\subsection*{Serotype proportion}
The susceptible population estimated in this paper is defined as individuals 
susceptible to at least one of the circulating serotypes in any given year. 
Let this Total number of susceptibles be denoted by $S_T$.

>>>>

% Our defs 
\def \rr {$R_{t}\ $}

@@

\subsection*{Estimating the reproductive number}

In monitoring of infectious diseases, it is important to assess whether the 
incidence of a a particular disease is increasing significantly, in order to 
decide to take preventive measures.
The effective reproductive number at time $t$, \rr, can be understood as a 
real-time estimate of the basic reproductive number ($R_{0}$) and is defined as 
the average number of secondary cases per primary case at time $t$.
Here we derive confidence/credibility bounds for the uncertainty about the 
estimate of the effective reproductive number $R_t$ under the assumption that 
reported disease counts are Poisson distributed.
It is shown that one can construct a conservative confidence interval for \rr 
using the induced probability distribution that results from the approach of 
\cite{mantel}.

Let $Y_t$ be the number of reported disease cases for a particular time $t \in 
(0, T)$.
\cite{nishiura} extend the theory developed by \cite{stallybrass} and propose 
to estimate \rr as
\begin{equation}
\label{eq:Rtestimate}
R_t = \left( \frac{Y_{t+1}}{Y_t}\right)^{1/n}
\end{equation}
where $n$ is taken to be the ratio between the length of reporting interval and 
the mean generation time of the disease.
Here we are interested in the simpler case $n=1$.
If \rr is to be used as a decision tool, one needs to be able to quantify the 
uncertainty about estimate in equation~\ref{eq:Rtestimate}. 
Here we detail how to obtain confidence intervals for \rr, exploring two 
approaches under the assumption that the counts $Y_t$ are Poisson distributed 
for all $t$.

First we explore the approach of \cite{mantel}, whose objective is to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $S = Y_{t} + Y_{t+1}$.
The authors then proceed to note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
\cite{mantel} show that one can construct a $100\alpha \%$ confidence interval 
for \rr by noting that
\begin{align}
 R_t^{(L)} &= \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})}\\
 R_t^{(U)} &= \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}
\end{align}

Many authors have derived $c_{\alpha}(\theta_t)$ following orthodox approaches 
such as those of \cite{wilson} and \cite{clopper}, mainly for simplicity 
reasons.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Equation~(\ref{eq:thetapost}) tells us that the induced distribution of $R_t$ is 
a beta prime with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 =  Y_t + b_0$.
The density of this induced distribution  $f_P(R_t)$ is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 - 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr under $f_P$ is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.

Since when $R_t > 1$ indicates sustained transmission, one may be interested in 
computing the probability of this event.
By noting that
\begin{align}
\label{cumprobMantel}
Pr(R_t > 1) &= 1 - \int_0^1 f_P(r)dr \\
            &= 1- Pr(\theta_t < \frac{1}{2})
\end{align}
one can avoid dealing with the density in~(\ref{eq:densityMantel}) directly.

The data used to fit the model consists of weekly incidence from 1996 to 2014 
in the city of Rio de Janeiro, Brazil.

@@

As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number as defined 
in~\ref{eq:Rtestimate}.
We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
 \frac{dS}{dt} &= -\beta(t)SI \\
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\
 \frac{dR}{dt} &= \tau I&
\end{align}
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
This is a rather simplified model, in which, for instance, the vector is left 
out.
Also, although there probably are multiple circulating serotypes, our approach
does not discriminate between them due to the lack of serotype-specific data.

@@

For the fitting procedure the time series was divided into $J=10$ epidemic 
windows that corresponded to significant raises in incidence and normalized to 
lie on the $[0,1]$ interval.
Also, as the numerical integration routine needs to treat $\beta(t)$ as a
continuous function of time because of the variable step size, we used linear 
interpolation to obtain values of $R_t$ for any time point.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techiniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at 
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubins' R convergence diagnostic~\cite{brooks1998}.

% Results and Discussion can be combined.
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\subsection*{Serotype proportion}
The susceptible population estimated in this paper is defined as individuals 
susceptible to at least one of the circulating serotypes in any given year. 
Let this Total number of susceptibles be denoted by $S_T$.
